---
title: "Exercise 2 - solutions: Advanced Tidyveres"
format: html
project:
 type: website
 output-dir: ../docs
---

## Introduction

In this exercise you will do some more advance tidyverse operations such as pivoting and nesting, as well as create plots to brush up on your ggplot skills.

## First steps

1.  Load packages.

```{r message=FALSE}
library(tidyverse)
```

2.  Load the joined diabetes data set you created in exercise 1 and the glucose dataset from the data folder.

```{r}
diabetes_join <- readxl::read_excel('../out/diabetes_join.xlsx')
df_glucose <- readxl::read_excel('../data/df_glucose.xlsx')
```

3. Have a look at the glucose dataset. The `OGTT` column contains measurements from a Oral Glucose Tolerance Test where blood glucose is measured at fasting (Glucose_0), 6 hours after glucose intake (Glucose_6), and 12 hours after (Glucose_12). 

```{r}
head(df_glucose)
```


4. Restructure the glucose dataset into a long format. How many rows are there per ID? Does that make sense?

::: {.callout-tip collapse="true"}
## Hint

Remember the flow:

```{r eval = FALSE}
pivot_longer(cols = LIST_WITH_COLUMNS_TO_PIVOT,
             names_to = "NEW_COLUMN_CONTAINING_COLUMN_NAMES",
             values_to = "NEW_COLUMN_CONTAINING_COLUMN_VALUES")
```

Have a look at slide 16 for a visual overview.

:::

```{r}
df_glucose_long <- df_glucose %>% 
  pivot_longer(cols = starts_with("Glucose"),
               names_to = "Measurement",
               values_to = "Glucose (mmol/L)"
               )

head(df_glucose_long)
```

5. Change the glucose measurements to numeric variable.

::: {.callout-tip collapse="true"}
## Hint
The `stringr` packages is a part of tidyverse and has many functions for manipulating strings. Find a function that can **split** the string so you can extract the numbers on the other side of the underscore. 
:::

```{r}
df_glucose_long <- df_glucose_long %>% 
  mutate(Measurement = str_split_i(Measurement, '_', 2) %>% as.numeric())

head(df_glucose_long)
```

6. Nest the glucose measurements and values such that there is only one row per ID.

::: {.callout-tip collapse="true"}
## Hint

Remember the flow:

```{r eval = FALSE}
group_by() %>% 
  nest() %>% 
  ungroup()
```

:::

```{r}
df_glucose_nested <- df_glucose_long %>% 
  group_by(ID) %>% 
  nest(OGTT = c(Measurement, `Glucose (mmol/L)`)) %>% 
  ungroup()
  
head(df_glucose_nested)
```


7. Merge the nested glucose dataset with the joined diabetes. 
```{r}
diabetes_glucose <- diabetes_join %>% 
  left_join(df_glucose_nested, by = 'ID')

head(diabetes_glucose)
```

8. Pull the glucose measurements (`OGTT`) from your favorite ID. 

::: {.callout-tip collapse="true"}
## Hint
First `filter` for your favorite ID and then `pull` the nested column. 
:::

```{r}
diabetes_glucose %>% 
  filter(ID == 9046) %>% 
  pull(OGTT)
```

9. Create a figure that visualizes glucose measurements at each time point (Measurement), stratified by patient ID. Give the plot a meaningful title.
```{r}
diabetes_glucose %>%
  unnest(OGTT) %>% 
  ggplot(aes(x = Measurement,
             y = `Glucose (mmol/L)`)) +
  geom_point(aes(color = Diabetes)) + 
  geom_line(aes(group = ID, color = Diabetes)) + 
  labs(title = 'Glucose Measurements Across Time Points by Diabetes Status')
```

10. Calculate the mean glucose measure for each measurement.

::: {.callout-tip collapse="true"}
## Hint
You will need to use `unnest()`, `group_by()`, and `summerize()`.
:::

```{r}
diabetes_glucose %>%
  unnest(OGTT) %>% 
  group_by(Measurement) %>% 
  summarize(`Glucose (mmol/L)` = mean(`Glucose (mmol/L)`))
```

11. Make the same calculation and stratify on Diabetes. 
```{r}
glucose_mean <- diabetes_glucose %>%
  unnest(OGTT) %>% 
  group_by(Measurement, Diabetes) %>%
  summarize(`Glucose (mmol/L)` = mean(`Glucose (mmol/L)`)) %>%
  ungroup()

glucose_mean
```

This next exercise might be a bit more challenging. It requires multiple operations and might involve some techniques that were not explicitly shown in the presentations. 

12. Recreate the plot you made in Exercise 10 and include the mean value for each glucose measurement for the two Diabetes statuses (0 and 1).

::: {.callout-tip collapse="true"}
## Hint
There are several ways to solve this task. There is a workflow suggestion:

  - The line in the plot is connected by ID. Create new IDs for the mean values that do not already exist in the dataset. Use `RANDOM_ID %in% df$ID` to check if an ID is already present. 
  
  - Data from another dataset can be added to the plot like this: `+ geom_point(DATA, aes(x = VAR1, y = VAR2, group = VAR3))`
  
:::

The points in the plot is connected by ID. Let's find two ID's (one for `Diabetes == 0` and another for `Diabetes == 1`) that are not in the data. We can use the same numbers as the Diabetes status. 
```{r}
0 %in% diabetes_glucose$ID
1 %in% diabetes_glucose$ID
```

Change the class of Measurement to factor and add ID to the glucose mean data set. 
```{r}
glucose_mean <- glucose_mean %>% 
  mutate(ID = Diabetes %>% as.double())

glucose_mean
```

Copy-paste the code in Exercise 9 and add lines with new data. 
```{r}
diabetes_glucose %>%
  unnest(OGTT) %>% 
  ggplot(aes(x = Measurement,
             y = `Glucose (mmol/L)`)) +
  geom_point(aes(color = Diabetes)) + 
  geom_line(aes(group = ID, color = Diabetes)) + 
  # Glucose mean data
  geom_point(data = glucose_mean, 
             aes(x = Measurement, 
                 y = `Glucose (mmol/L)`, 
                 group = ID)) + 
  geom_line(data = glucose_mean, 
            aes(x = Measurement, 
                y = `Glucose (mmol/L)`, 
                group = ID, 
                linetype = Diabetes)) + 
  labs(title = "Glucose Measurements with Mean by Diabetes Status")

ggsave('../out/figure2_12.png')
```

13. Export the final dataset. Since the dataset is nested, you cannot export it as an excel file. Export the dataset as an `.rds` file. Have a guess at what the function is called. 
```{r}
write_rds(diabetes_glucose, '../out/diabetes_glucose.rds')
```

