---
title: "Exercise 4 - Solutions"
format: html
project:
 type: website
 output-dir: ../docs
---

## Functions

1. Create a file that will contain your functions. 

2. In the file, write a function that takes a dataframe and a numerical value and returns a boxplot. Do not worry about error or warning handling yet. **NB** If you modify the function script, be sure to save it and `source` it again so that your updates are reflected in the main script.

::: {.callout-tip collapse="true"}
## Hint

`aes_string` allows the function to dynamically reference a column name from the dataframe.

Example: `aes_string("colname")`

:::

## Main script

3. Create a Quarto document and start by loading relevant packages. 
```{r message=FALSE}
library(tidyverse)
library(glue)
```

4. Source the function script into the Quarto document. 
```{r}
source('solution4_functions.R')
```

5. Load data from the `.rds` you created in Exercise 2. 
```{r}
diabetes_glucose <- read_rds('../out/diabetes_glucose.rds')
```

6. Run the boxplot function for the diabetes dataset and the `Age` variable.
```{r}
boxplot(df = diabetes_glucose, var_num = 'Age')
```

7. In a for-loop, run the boxplot function for the diabetes dataset and each of the numerical variables in the script. 
```{r}
numerical_columns <- sapply(diabetes_glucose, is.numeric)
numerical_colnames <- names(diabetes_glucose)[numerical_columns]
```

```{r}
for (var_num in numerical_colnames){
  
  p <- boxplot(df = diabetes_glucose, var_num = var_num)
  print(p)
  # ggsave(paste0("../out/boxplot_", var_num,".png"))
  
}
```

8. Go back to your function script. Create a function that takes a dataframe and a categorical value and returns a bar plot.

9. Run the function on the `Sex` variable. 
```{r}
bar_plot(df = diabetes_glucose, var_cat = 'Sex')
```

10. In a for-loop, run the bar plot function for the diabetes dataset and each of the categorical variables in the script.
```{r}
categorical_columns <- sapply(diabetes_glucose, is.character)
categorical_colnames <- names(diabetes_glucose)[categorical_columns]
```

```{r}
for (var_cat in categorical_colnames){
  
  p <- bar_plot(df = diabetes_glucose, var_cat = var_cat)
  print(p)
  # ggsave(paste0("../out/bar_plot_", var_cat,".png"))
  
}
```

11. It's time for a fun exercise! Customize your plot functions by adding a theme, colors and labels. Add a `ggsave` statement in your for-loops to recreate and save your plots.

## Functions and error handling 

12. In your function script, copy the function you created in Exercise 2 and modify it so that it generates an error if `var_num` is assigned a non-numerical argument. 

::: {.callout-tip collapse="true"}
## Hint

When checking the class of the variables, make sure that they are in vector format. Have a look at Presentation 1 if you are unsure of how to access a variable in a data frame as a vector. 

```{r eval=FALSE}
is.numeric(vector)
```

:::

13. Test the function by using `Sex` as the numerical variable argument.
```{r eval=FALSE}
boxplot_2(df = diabetes_glucose, var_num = 'Sex')
# Gives an error.
```

14. Use the same logic for the function you created in Exercise 8.

15. Test the function by using `Age` as the categorical variable argument.
```{r eval=FALSE}
bar_plot_2(df = diabetes_glucose, var_cat = 'Age')
# Gives an error.
```

16. In Exercise 7, a boxplot have made for the `ID` variable if you simply looped over all the numcerical values. Avoid this using an if-statement. 
```{r}
for (var_num in numerical_colnames){
  
  if (var_num != 'ID'){
    
    p <- boxplot(df = diabetes_glucose, var_num = var_num)
    print(p)
    
  }
  
}
```

17. Make a function that takes an ID and outputs a graph of the glucose measurement over time. There are several way to do this depending on how generalized your function is. The more generalized, the more difficult. 

::: {.callout-tip collapse="true"}
## Hint

Remember that you need to put backticks when you access a `colname with spaces`. 

When using `aes_string`, you need backticks **and** quotation marks: ``aes_string("`colname with spaces`")``. 

:::

18. Run the function on your favorite ID. 
```{r}
df <- diabetes_glucose %>% unnest(OGTT)

glucose_measurement_id_plot(df = df, id = "12109")
```

19. Inspired by Exercise 3.13, create a function that takes a dataframe and a categorical variable as input, calculates the mean glucose levels over time for each category, and produces a line plot showing the mean glucose measurements across time points.

::: {.callout-tip collapse="true"}
## Hint

The function requires the column name to be passed as a string. However, not all tidyverse functions can directly handle column names in this format. Fortunately, there is a workaround. For example, the `group_by()` function can dynamically reference a column name provided as a string using `.data[[...]]`:
```{r eval=FALSE}
var_cat <- "Age"

df %>% 
    group_by(Measurement, .data[[var_cat]])
```

Here, `.data` refers to the data frame being used — in this case, `df.`

Also, you might get a warning like this *You can override using the* `.groups` *argument.* This can be solved adding `.groups = "drop"` to the `summerize` function. 

:::

20. Run the function with `Sex` as input. 
```{r}
glucose_measurement_mean_plot(df = df, var_cat = 'Sex')
```

21. In a for-loop, run the function on all categorical variables in the data frame. Between which groups do you visually notice the largest differences in the mean OGTT?
```{r}
for (var_cat in categorical_colnames){
  
  p <- glucose_measurement_mean_plot(df = df, var_cat = var_cat)
  print(p)
  
}
```


Last. Render you document and review the generated html. If you get errors, correct them and re-render. 
