---
title: "Presentation 1 - Solutions: Base R and Tidyverse / Data Clean-up and Wrangling"
format: html
project:
 type: website
 output-dir: ../docs
---

## Getting started

1.  Load packages.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

2.  Load in the `diabetes_clinical_toy_messy.xlsx` data set.

```{r}
diabetes_clinical <- read_excel('../data/diabetes_clinical_toy_messy.xlsx')
head(diabetes_clinical)
```

## Explore the data

Use can you either base R or/and tidyverse to solve the exercises. 

3.  How many missing values (NA's) are there in each column. 

```{r}
colSums(is.na(diabetes_clinical))
```

4.  Check the distribution of each of the variables. Consider that the variables are of different classes. Do any of the distributions seem odd to you?

For the categorical variables:

The `Sex` values are not consistent. 
```{r}
table(diabetes_clinical$Sex)
table(diabetes_clinical$Smoker)
table(diabetes_clinical$Diabetes)
```

For the numerical variables:

```{r}
diabetes_clinical %>% 
  ggplot(aes(y = Age)) + 
  geom_boxplot()
```

Odd: Some `BloodPressure` values are 0. 
```{r}
diabetes_clinical %>% 
  ggplot(aes(y = BloodPressure)) + 
  geom_boxplot()
```

```{r}
diabetes_clinical %>% 
  ggplot(aes(y = GeneticRisk)) + 
  geom_boxplot()
```
Odd: Some `BMI` values are 0. 
```{r}
diabetes_clinical %>% 
  ggplot(aes(y = BMI)) + 
  geom_boxplot()
```

```{r}
diabetes_clinical %>% 
  ggplot(aes(y = PhysicalActivity)) + 
  geom_boxplot()
```

## Clean up the data

Now that we have had a look at the data, it is time to correct fixable mistakes and remove observations that cannot be corrected.

Consider the following:

-   What should we do with the rows that contain NA's? Do we remove them or keep them?

-   Which odd things in the data can we correct with confidence and which cannot?

-   Are there zeros in the data? Are they true zeros or errors?

-   Do you want to change any of the classes of the variables?

5.  Clean the data according to your considerations.

::: {.callout-tip collapse="true"}
## Hint
Have a look at `BloodPressure`, `BMI`, `Sex`, and `Diabetes`. 
:::

My considerations:

-   When modelling, rows with NA's in the variables we want to model should be removed as we cannot model on NAs. Since there are only NA's in `GeneticRisk`, the rows can be left until we need to do a model with `GeneticRisk.`

-   The uppercase/lowercase mistakes in `Sex` does not influence the interpretability of the variables, so they are simply changes such that the first letter is a capital letter and the remaining letter are lowercase.

-   There are zeros in `BMI` and `BloodPressure`. These are considered false zeros as is does not make sense that these variables have a value of 0.

-   `Diabetes` is changed to factor. 

Access number of rows before cleaning.
```{r}
nrow(diabetes_clinical)
```

Cleaning data according to considerations.
```{r}
diabetes_clinical_clean <- diabetes_clinical %>% 
  mutate(Sex = str_to_title(Sex),
         Diabetes = Diabetes %>% factor()) %>% 
  filter(BMI != 0, BloodPressure != 0) 
```

Check the unique sexes now.
```{r}
diabetes_clinical_clean$Sex %>% unique()
```

Access number of rows after cleaning. 
```{r}
nrow(diabetes_clinical_clean)
```

## Meta Data

6.  Perform step 2-5 from above and do data exploration and cleaning workflow for the `diabetes_meta_toy_messy.csv` data set. Use the `read_delim` function to load in the dataset. 

6.2. Load the data set.

```{r message=FALSE}
diabetes_meta <- read_delim('../data/diabetes_meta_toy_messy.csv')
head(diabetes_meta)
```

Use can you either base R or/and tidyverse to solve the exercises. For now, we just explore the data.

6.3. How many missing values (NA's) are there in each column. 

```{r}
colSums(is.na(diabetes_meta))
```

6.4. Check the distribution of each of the variables. Consider that the variables are of different classes. Do any of the distributions seam odd to you?

For the categorical variables:
```{r}
table(diabetes_meta$Married)
table(diabetes_meta$Work)
```

By investigating the unique values of the `Married` variable we see that some of the values have whitespace. 
```{r}
unique(diabetes_meta$Married)
```

6. 5.  Clean the data according to your considerations.

My considerations:

-   The `Married` variable has whitespace in the some of the values. The values "Yes" and "Yes " will be interpreted as different values. We can confidently remove all the whitespaces in this variable. 

-   `ID` is changed to numerical to match the `diabetes_clean` dataset. 

Access number of rows before cleaning.
```{r}
nrow(diabetes_meta)
```

```{r}
diabetes_meta_clean <- diabetes_meta %>% 
  mutate(Married = str_trim(Married),
         ID = as.numeric(ID))
```

Check the unique marital status now.
```{r}
unique(diabetes_meta_clean$Married)
```

Access number of rows after cleaning.
```{r}
nrow(diabetes_meta_clean)
```

## Join the datasets

7. Consider what variable the datasets should be joined on.

::: {.callout-tip collapse="true"}
## Hint
The joining variable must be the same type in both datasets.
:::

8. Join the datasets by the variable you selected above.
```{r}
diabetes_join <- diabetes_clinical_clean %>% 
  left_join(diabetes_meta_clean, by = 'ID')
```

9. How many rows does the joined dataset have? Explain why. 

::: {.callout-tip collapse="true"}
## Hint
Because we used `left_join`, only the IDs that are in `diabetes_clinical_clean` are kept. 
:::

```{r}
nrow(diabetes_join)
```

10. Export the joined dataset. Think about which directory you want to save the file in. 
```{r}
writexl::write_xlsx(diabetes_join, '../out/diabetes_join.xlsx')
```

