---
title: "Untitled"
format: html
editor: visual
---
## Regression with categorical features

Now that we know how to make a simple linear model, how can we include categorical variables and what is the interpretation of their coefficients? To investigate this we include the other predictor variable we have: Exercise level.

```{r}
distinct(df_smoke, exercise)
```

Alright, we have three different levels of exercise. They are: low == 0, moderate == 1 and high == 2. Before we go on, let's have a look if our data is represented correctly:

```{r}
summary(df_smoke)
```

We can see that the `exercise` column is interpreted as an integer. However, it is actually a category! In R categorical variables are known as `factors` and have their own datatype. Let's convert exercise to a factor:

```{r}
df_smoke$exercise <- as.factor(df_smoke$exercise)

df_smoke$exercise
```

As before, before fitting the model we'll split up the data in train and test. Since we're using the same seed we should get the same observations, i.e. rows into training and test as above.

```{r}
# Set seed to ensure reproducibility
set.seed(123)  

#add an ID column to keep track of observations
df_smoke$ID <- 1:nrow(df_smoke)

train <- df_smoke %>% sample_frac(.75)
test  <- anti_join(df_smoke, train, by = 'ID') 
```

And now we extend our previous model formula with the new predictor:

```{r}
model2 <- lm(life ~ daily_cigarettes + exercise, data = train)
```

```{r}
summary(model2)
```



When we check the summary we see that it has two additional coefficients, `exercise1` and `exercise2`. What are they?

Because `exercise` is a categorical variable it is dummy coded. That means our model formula mathematically looks something like this:

$$y=b_0+b_1*x_1 + b_2 *x_2 + b_3*x_3$$

with:

| Exercise level | $x_2$ | $x_3$ |
|:--------------:|:-----:|------:|
|       0        |   0   |     0 |
|       1        |   1   |     0 |
|       2        |   0   |     1 |

And for our coefficients it means:

```{r}
model2$coefficients
```

-   `Intercept` == $b_0$: The life expectancy at 0 cigarettes and exercise level 0

-   `daily_cigerettes` == $b_1$: The change in life expectancy for each additional cigarette.

-   `exercise1` == $b_2$: The change in life expectancy if the exercise level is 1 (assuming the number of cigarettes stays constant).

-   `exercise2` == $b_3$: The change in life expectancy if the exercise level is 2 (assuming the number of cigarettes stays constant).

Why is there no coefficient for exercise level 0 (low amount of exercise)? This case is covered in the Intercept. It is referred to as the **reference level** of the categorical variable. You can change which level is regarded as the reference and the effect of having this level will always be modelled in the intercept.

## C